---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  eval = TRUE
)

library(dplyr)
library(pryr)
library(ggplot2)
library(bench)
library(tidyr)
library(patchwork)

library(insitu)
```

# insitu

<!-- badges: start -->
![](https://img.shields.io/badge/cool-useless-green.svg)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
<!-- badges: end -->

`insitu` provides some functions for modifying vectors *in-situ* (in place) 
without allocating a new vector.

This is an experimental package I wrote to try and figure out just how fast things 
could be in R.

**Warning**: Modifying vectors in-situ can be fast, but doing so violates some very strong norms
within R internals.  

**Warning**: Unless you are very confident on who has references to your data,
do not use this package for in-place modification.

## Modifying vectors in-place


**Pros** of modifying in-situ

* Often faster as there is reduced memory allocation (and also reduced pressure
  on the garbage collector)
* Often faster as there is no copying from the original vector into a new vector
* By using C, we can override the *copy-on-modify* semantics usually used in R
  (regardless of how many references exist to the given object)


**Cons** of modifying in-situ

* Normal R *copy-on-modify* behaviour is not followed - this will be a point of
  confusion as essentially everything in R uses *copy-on-modify* and not 
  modification in-place.
* Copying and allocating vectors is already very fast.  It only takes a 
  few *microseconds* to allocate memory for a vector of 1000 elements and assign new values into 
  it.  The speed saved by switching to in-situ modification will only possibly
  be useful if this operations is performed many, many times.



In the words of Luke Tierney in his User2020 keynote:

    You should never modify something without duplicating it.

This package ignores this advice. Beware.




## What's in the box

| insitu                                  | description                                                            | integer | real | character |
|-----------------------------------------|------------------------------------------------------------------------|---------|------|-----------|
| insitu_fill(x, value)                   | Fill vector with the given value                                       | Yes     | Yes  | Yes       |
| insitu_fill_runif(x, lower, upper)      | Fill vector with uniform random numbers                                | Yes     | Yes  |           |
| insitu_fill_runif_fast(x, lower, upper) | Fill vector with uniform random numbers                                | Yes     | Yes  |           |
| insitu_replace(x, pos, values)          | Replace values in x with given values starting from the given position | Yes     | Yes  | Yes       |
| insitu_reverse(x)                       | Reverse vector                                                         | Yes     | Yes  | Yes       |
| insitu_shuffle(x)                       | Shuffle the elements of a vector                                       | Yes     | Yes  | Yes       |
| insitu_shuffle_fast(x)                  | Shuffle the elements of a vector                                       | Yes     | Yes  | Yes       |
| insitu_sort(x)                          | Sort the elements of a vector                                          | Yes     | Yes  |           |
|                                         |                                                                        |         |      |           |

#### Fast Variants

The `_fast` versions of some functions use their own random-number generator 
rather than the one supplied in R.  This generator is [lehmer64](https://lemire.me/blog/2019/03/19/the-fastest-conventional-random-number-generator-that-can-pass-big-crush/).
It is very fast, but has different properties from R's built in random 
number generator. Use with caution.

#### ALTREP utils

* `is_altrep(x)` tests whether an object is an ALTREP
* `is_mutable(x)` tests whether an object is mutable in-place by checking its
  reference count
* `get_refcnt(x)` returns the reference count for the object 



## Installation

You can install from [GitHub](https://github.com/coolbutuseless/insitu) with:

``` r
# install.package('remotes')
remotes::install_github('coolbutuseless/insitu')
```


## Future possibilities

* In-place radix sort - instead of using `qsort()`
* `cummin`, `cummax` etc
* `negate`
* `diff`


## In-situ Replacement

`insitu_replace()` is analogous to `replace()` but replaces values in the current 
vector rather than creating a new one.

```{r}
x <- integer(10)
insitu_replace(x, 6, 1:5)
x
```




<details>
<summary> Click to show/hide benchmark code & results </summary>

```{r message=FALSE}
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Assign values into a vector
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
N = c(1e1, 1e2, 1e3, 1e4, 1e5, 1e6)

res_rel <- bench::press(
  N = N,
  {
    x <- numeric(N)
    x[1] <- x[1]
    idx <- seq(1, N/2)
    ins <- as.numeric(idx)
    
    bench::mark(
      replace(x, idx, ins),
      insitu_replace(x, 1, ins),
      check    = TRUE,
      relative = TRUE
    )
  }
)

res_rel %>%
  select(expression, min, median, `itr/sec`, mem_alloc) %>%
  knitr::kable()

res_abs <- bench::press(
  N = N,
  {
    x <- numeric(N)
    x[1] <- x[1]
    idx <- seq(1, N/2)
    ins <- as.numeric(idx)
    
    bench::mark(
      replace(x, idx, ins),
      insitu_replace(x, 1, ins),
      check    = TRUE,
      relative = FALSE
    )
  }
)

res_abs %>%
  select(expression, min, median, `itr/sec`, mem_alloc) %>%
  knitr::kable()
```


</details>

```{r echo=FALSE}

p_rel <- ggplot(res_rel, aes(N, min,colour=as.character(expression))) +
  geom_point() + 
  geom_line(aes(group = as.character(expression))) +
  theme_bw() + 
  scale_x_log10(breaks = N) + 
  theme(legend.position = 'bottom') + 
  labs(
    x        = "Vector Length",
    y        = "Relative duration compared to fastest method"
  )


p_abs <- ggplot(res_abs, aes(N, min,colour=as.character(expression))) +
  geom_point() + 
  geom_line(aes(group = as.character(expression))) +
  theme_bw() + 
  scale_x_log10(breaks = N) + 
  theme(legend.position = 'bottom') + 
  labs(
    x        = "Vector Length",
    y        = "Minimum time for call"
  )

p <- p_rel + p_abs + plot_annotation(
  title = "in-situ assignment vs copy-on-modify assignment",
  subtitle = "Assign N/2 values at the start of a vector of length N\nLower is better",
  theme = theme(legend.position = 'bottom')
) + 
  plot_layout(guides = 'collect') 

p
```




## In-situ fill


`insitu_fill()` is analogous to `replace()` but assigns the value into the current 
vector rather than creating a new one.


```{r}
x <- integer(10)
insitu_fill(x, 3L)
x
```

<details>
<summary> Click to show/hide benchmark code & results </summary>

```{r message=FALSE}
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Assign values into a vector
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# N = c(1e1, 1e2, 1e3, 1e4, 1e5, 1e6)

res_rel <- bench::press(
  N = N,
  {
    x <- numeric(N)
    idx <- 1:N
    ins <- 99
    
    bench::mark(
      replace(x, idx, ins),
      insitu_fill(x, ins),
      check    = TRUE,
      relative = TRUE
    )
  }
)

res_rel %>%
  select(expression, min, median, `itr/sec`, mem_alloc) %>%
  knitr::kable()

res_abs <- bench::press(
  N = N,
  {
    x <- numeric(N)
    idx <- 1:N
    ins <- 99
    
    bench::mark(
      replace(x, idx, ins),
      insitu_fill(x, ins),
      check    = TRUE,
      relative = FALSE
    )
  }
)

res_abs %>%
  select(expression, min, median, `itr/sec`, mem_alloc) %>%
  knitr::kable()
```


</details>

```{r echo=FALSE}

p_rel <- ggplot(res_rel, aes(N, min,colour=as.character(expression))) +
  geom_point() + 
  geom_line(aes(group = as.character(expression))) +
  theme_bw() + 
  scale_x_log10(breaks = N) + 
  theme(legend.position = 'bottom') + 
  labs(
    x        = "Vector Length",
    y        = "Relative duration compared to fastest method"
  )


p_abs <- ggplot(res_abs, aes(N, min,colour=as.character(expression))) +
  geom_point() + 
  geom_line(aes(group = as.character(expression))) +
  theme_bw() + 
  scale_x_log10(breaks = N) + 
  theme(legend.position = 'bottom') + 
  labs(
    x        = "Vector Length",
    y        = "Minimum time for call"
  )

p <- p_rel + p_abs + plot_annotation(
  title = "Fill: in-situ vs copy-on-modify",
  subtitle = "Fill a vector with a given single value\nLower is better",
  theme = theme(legend.position = 'bottom')
) + 
  plot_layout(guides = 'collect') 

p
```




## In-situ sort

`insitu_sort()` is analogous to `sort()` but sorts values in the current 
vector rather than creating a new one.

```{r}
x <- sample(10)
insitu_sort(x)
x
```

<details>
<summary> Click to show/hide benchmark code & results </summary>

```{r message=FALSE}
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Assign values into a vector
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# N = c(1e1, 1e2, 1e3, 1e4, 1e5, 1e6)

res_rel <- bench::press(
  N = N,
  {
    x <- runif(N)

    bench::mark(
      sort(x),
      insitu_sort(x),
      check    = TRUE,
      relative = TRUE
    )
  }
)

res_rel %>%
  select(expression, min, median, `itr/sec`, mem_alloc) %>%
  knitr::kable()

res_abs <- bench::press(
  N = N,
  {
    x <- runif(N)
    
    bench::mark(
      sort(x),
      insitu_sort(x),
      check    = TRUE,
      relative = FALSE
    )
  }
)

res_abs %>%
  select(expression, min, median, `itr/sec`, mem_alloc) %>%
  knitr::kable()
```


</details>

```{r echo=FALSE}

p_rel <- ggplot(res_rel, aes(N, min,colour=as.character(expression))) +
  geom_point() + 
  geom_line(aes(group = as.character(expression))) +
  theme_bw() + 
  scale_x_log10(breaks = N) + 
  theme(legend.position = 'bottom') + 
  labs(
    x        = "Vector Length",
    y        = "Relative duration compared to fastest method"
  )


p_abs <- ggplot(res_abs, aes(N, min,colour=as.character(expression))) +
  geom_point() + 
  geom_line(aes(group = as.character(expression))) +
  theme_bw() + 
  scale_x_log10(breaks = N) + 
  theme(legend.position = 'bottom') + 
  labs(
    x        = "Vector Length",
    y        = "Minimum time for call"
  )

p <- p_rel + p_abs + plot_annotation(
  title = "Sort: in-situ vs copy-on-modify",
  subtitle = "Sort a vector\nLower is better",
  theme = theme(legend.position = 'bottom')
) + 
  plot_layout(guides = 'collect') 

p
```



## In-situ Shuffle


`insitu_shuffle()` is analogous to `sample()` but shuffles values in the current 
vector rather than creating a new one.

```{r}
x <- c(0, 1, 2, 3, 4, 5, 6, 7, 8, 9)
insitu_shuffle(x)
x
```


<details>
<summary> Click to show/hide benchmark code & results </summary>

```{r message=FALSE}
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Assign values into a vector
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# N = c(1e1, 1e2, 1e3, 1e4, 1e5, 1e6)

res_rel <- bench::press(
  N = N,
  {
    x <- runif(N)

    bench::mark(
      sample(x),
      insitu_shuffle(x),
      insitu_shuffle_fast(x),
      check    = FALSE,
      relative = TRUE
    )
  }
)

res_rel %>%
  select(expression, min, median, `itr/sec`, mem_alloc) %>%
  knitr::kable()

res_abs <- bench::press(
  N = N,
  {
    x <- runif(N)
    
    bench::mark(
      sample(x),
      insitu_shuffle(x),
      insitu_shuffle_fast(x),
      check    = FALSE,
      relative = FALSE
    )
  }
)

res_abs %>%
  select(expression, min, median, `itr/sec`, mem_alloc) %>%
  knitr::kable()
```


</details>

```{r echo=FALSE}

p_rel <- ggplot(res_rel, aes(N, min,colour=as.character(expression))) +
  geom_point() + 
  geom_line(aes(group = as.character(expression))) +
  theme_bw() + 
  scale_x_log10(breaks = N) + 
  theme(legend.position = 'bottom') + 
  labs(
    x        = "Vector Length",
    y        = "Relative duration compared to fastest method"
  )


p_abs <- ggplot(res_abs, aes(N, min,colour=as.character(expression))) +
  geom_point() + 
  geom_line(aes(group = as.character(expression))) +
  theme_bw() + 
  scale_x_log10(breaks = N) + 
  theme(legend.position = 'bottom') + 
  labs(
    x        = "Vector Length",
    y        = "Minimum time for call"
  )

p <- p_rel + p_abs + plot_annotation(
  title = "Shuffle: in-situ vs copy-on-modify",
  subtitle = "Randomly shuffle a vector\nLower is better",
  theme = theme(legend.position = 'bottom')
) + 
  plot_layout(guides = 'collect') 

p
```




## In-situ Reverse

`insitu_reverse()` is analogous to `rev()` but reverses values in the current 
vector rather than creating a new one.


```{r}
x <- c(0, 1, 2, 3, 4, 5, 6, 7, 8, 9)
insitu_reverse(x)
x
```


<details>
<summary> Click to show/hide benchmark code & results </summary>

```{r message=FALSE}
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Assign values into a vector
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# N = c(1e1, 1e2, 1e3, 1e4, 1e5, 1e6)

res_rel <- bench::press(
  N = N,
  {
    x <- runif(N)

    bench::mark(
      rev(x),
      insitu_reverse(x),
      check    = TRUE,
      relative = TRUE
    )
  }
)

res_rel %>%
  select(expression, min, median, `itr/sec`, mem_alloc) %>%
  knitr::kable()

res_abs <- bench::press(
  N = N,
  {
    x <- runif(N)
    
    bench::mark(
      rev(x),
      insitu_reverse(x),
      check    = TRUE,
      relative = FALSE
    )
  }
)

res_abs %>%
  select(expression, min, median, `itr/sec`, mem_alloc) %>%
  knitr::kable()
```


</details>

```{r echo=FALSE}

p_rel <- ggplot(res_rel, aes(N, min,colour=as.character(expression))) +
  geom_point() + 
  geom_line(aes(group = as.character(expression))) +
  theme_bw() + 
  scale_x_log10(breaks = N) + 
  theme(legend.position = 'bottom') + 
  labs(
    x        = "Vector Length",
    y        = "Relative duration compared to fastest method"
  )


p_abs <- ggplot(res_abs, aes(N, min,colour=as.character(expression))) +
  geom_point() + 
  geom_line(aes(group = as.character(expression))) +
  theme_bw() + 
  scale_x_log10(breaks = N) + 
  theme(legend.position = 'bottom') + 
  labs(
    x        = "Vector Length",
    y        = "Minimum time for call"
  )

p <- p_rel + p_abs + plot_annotation(
  title = "Reverse: in-situ vs copy-on-modify",
  subtitle = "Reverse a vector\nLower is better",
  theme = theme(legend.position = 'bottom')
) + 
  plot_layout(guides = 'collect') 

p
```







## In-situ fill with random

`insitu_fill_runif()` is analogous to `runif()` but generates values in the current 
vector rather than creating a new one.


```{r}
x <- integer(10)
insitu_fill_runif(x, 10, 15)
x
```

<details>
<summary> Click to show/hide benchmark code & results </summary>

```{r message=FALSE}
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Assign values into a vector
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# N = c(1e1, 1e2, 1e3, 1e4, 1e5, 1e6)

res_rel <- bench::press(
  N = N,
  {
    x <- numeric(N)

    bench::mark(
      runif(x),
      insitu_fill_runif(x, 0, 1),
      insitu_fill_runif_fast(x, 0, 1),
      check    = FALSE,
      relative = TRUE
    )
  }
)

res_rel %>%
  select(expression, min, median, `itr/sec`, mem_alloc) %>%
  knitr::kable()

res_abs <- bench::press(
  N = N,
  {
    x <- numeric(N)
    
    bench::mark(
      runif(x),
      insitu_fill_runif(x, 0, 1),
      insitu_fill_runif_fast(x, 0, 1),
      check    = FALSE,
      relative = FALSE
    )
  }
)

res_abs %>%
  select(expression, min, median, `itr/sec`, mem_alloc) %>%
  knitr::kable()
```


</details>

```{r echo=FALSE}

p_rel <- ggplot(res_rel, aes(N, min,colour=as.character(expression))) +
  geom_point() + 
  geom_line(aes(group = as.character(expression))) +
  theme_bw() + 
  scale_x_log10(breaks = N) + 
  theme(legend.position = 'bottom') + 
  labs(
    x        = "Vector Length",
    y        = "Relative duration compared to fastest method"
  )


p_abs <- ggplot(res_abs, aes(N, min,colour=as.character(expression))) +
  geom_point() + 
  geom_line(aes(group = as.character(expression))) +
  theme_bw() + 
  scale_x_log10(breaks = N) + 
  theme(legend.position = 'bottom') + 
  labs(
    x        = "Vector Length",
    y        = "Minimum time for call"
  )

p <- p_rel + p_abs + plot_annotation(
  title = "Random Fill: in-situ vs copy-on-modify",
  subtitle = "Fill a vector with uniform random values\nLower is better",
  theme = theme(legend.position = 'bottom')
) + 
  plot_layout(guides = 'collect') 

p
```




## Related Software

* [data.table](https://cran.r-project.org/package=data.table) performs a lot of 
  operations in-situ (which data.table calls "by reference")


## Acknowledgements

* R Core for developing and maintaining the language.
* CRAN maintainers, for patiently shepherding packages onto CRAN and maintaining
  the repository
